\documentclass[
paper=a4,                       % paper size
fontsize=11pt,                  % font size
twoside,                        % two sided
footsepline,                    % add a line to separate the footer
headsepline,                    % add a line to separate the header
headinclude=false,              % header does not belong to the text
footinclude=false,              % footer does not belong to the text
pagesize,                       % set the pagesize in a DVI document
]{scrartcl}

\include{common}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{listings}
\lstset{language=Python}

\newtheorem{task}{Task}

\begin{document}

\esptitlehead

\title{Tutorial 8: Visualization%
\ifdefined\esversion%
\thanks{For \es \esversion}%
\fi%
}
\subtitle{How to visualize your \es simulations while they are running}
\author{M. Kuron, K. Breitsprecher}
\maketitle

\section{Introduction}
\label{intro}

When you are running a simulation, it is often useful to see what is going on by visualizing particles in a 3D view or by plotting observables over time.
That way, you can easily determine things like whether your choice of parameters has led to a stable simulation or whether your system has equilibrated.
You may even be able to do your complete data analysis in real time as the simulation progresses.\\

\noindent Thanks to \es's Python interface, we can make use of standard libraries like Mayavi or OpenGL (for interactive 3D views) and Matplotlib (for line graphs) for this purpose.
We will also use NumPy, which both of these libraries depend on, to store data and perform some basic analysis.

\section{Simulation}
\label{sim}

First, we need to set up a simulation.
We simulate a simple Lennard-Jones liquid in this tutorial using the script shown below.
Visualization of more advanced features of \es is also possible (for example, bonds) or will be possible soon (for example, constraints and lattice Boltzmann).

\lstinputlisting{scripts/simulation.py}

\section{Live plotting}
\label{plot}

Let's have a look at the total energy of the simulation.
We can determine the individual energies in the system using
\begin{lstlisting}
print (system.analysis.energy())
\end{lstlisting}
to get
\begin{lstlisting}
OrderedDict([
        ('total', 1840.118038871784),
        ('ideal', 1358.743742464325),
        ('bonded', 0.0), 
        (('nonBonded', 0, 0), 481.374296407459), 
        ('nonBonded', 481.374296407459), 
        ('coulomb', 0.0)])
\end{lstlisting}
Write that command right after the call to \lstinline{main()} and see if you can get a similar result.
Now we want to store the total energy over time in a NumPy array.
To do that, modify the \lstinline{main} function definition to be the following:
\begin{lstlisting}
energies = numpy.empty((int_steps,2))
def main():
    for i in range(0, int_n_times):
        print("run %d at time=%f " % (i, system.time))
        system.integrator.run(int_steps)
        energies[i] = (system.time, system.analysis.energy()['total'])
\end{lstlisting}
Now we can do some analysis on the stored energies.
For example, let us calculate the time-averaged energy:
\begin{lstlisting}
print ("Average energy: %.6g" % energies[:,1].mean())
\end{lstlisting}
We can also plot the energy over time by adding
\begin{lstlisting}
pyplot.xlabel("time")
pyplot.ylabel("energy")
pyplot.plot(energies[:,0],energies[:,1])
pyplot.show()
\end{lstlisting}
to the end of the script. Of course, this plot only gets shown after the entire simulation is completed.
To get an interactive plot, we update it from within the integration loop.
\begin{lstlisting}
energies = numpy.empty((int_steps,2))
current_time = -1
pyplot.xlabel("time")
pyplot.ylabel("energy")
plot, = pyplot.plot([0],[0])
pyplot.show(block=False)
def update_plot():
    if current_time < 0:
        return
    i = current_time
    plot.set_xdata(energies[:i+1,0])
    plot.set_ydata(energies[:i+1,1])
    pyplot.xlim(0, energies[i,0])
    pyplot.ylim(energies[:i+1,1].min(), energies[:i+1,1].max())
    pyplot.draw()
def main():
    global current_time
    for i in range(0, int_n_times):
        print("run %d at time=%f " % (i, system.time))
        system.integrator.run(int_steps)
        energies[i] = (system.time, system.analysis.energy()['total'])
        current_time = i
        update_plot()
\end{lstlisting}

\noindent One shortcoming of this simple method is that one cannot interact with the controls of the plot window (e.g. resize the window or zoom around in the graph).
This will be resolved using multiple threads when we combine the plotting with the 3D visualization in the next section.

\section{Live visualization}
\label{vis}

In order to be able to interact with the live visualization, we need to move the main integration loop into a secondary thread and run the visualization in the main thread (note that visualization or plotting cannot be run in secondary threads). First, let's revert to the main loop without plotting:
\begin{lstlisting}
energies = numpy.empty((int_steps,2))
def main():
    for i in range(0, int_n_times):
        print("run %d at time=%f " % (i, system.time))
        system.integrator.run(int_steps)
        energies[i] = (system.time, system.analysis.energy()['total'])
\end{lstlisting}

Then, add the following line after the particle setup code:
\begin{lstlisting}
visualizer = visualization.mayaviLive(system)
\end{lstlisting}
Now, go to the end of the \lstinline{main} function definition and add
\begin{lstlisting}
visualizer.update()
\end{lstlisting}
which sends the current simulation state to the visualizer.
Now, go to the line where \lstinline{main()} is called and replace it with the following code, which dispatches the function in a secondary thread, and then opens the visualizer window:
\begin{lstlisting}
t = Thread(target=main)
t.daemon = True
t.start()
visualizer.start()
\end{lstlisting}
While the simulation is running, you can move and zoom around with your mouse or explore the buttons in the toolbar to see how the graphical representation can be changed.\\

\noindent Alternatively, you can try \es's own OpenGL visualization by switching the visualizer to:
\begin{lstlisting}
visualizer = visualization.openGLLive(system)
\end{lstlisting}
To follow the trajectories, try decreasing the integration steps to 1 and the time step to 0.001. 
You may have noticed that you can drag particles with your mouse. 

\section{Combined live visualization and plotting}

Now let's merge the code from the preceding two sections so we can see the energy graph while viewing the 3D visualization of the particles.
Do do that, we copy the \lstinline{pyplot}-related lines from above:
\begin{lstlisting}
current_time = -1
pyplot.xlabel("time")
pyplot.ylabel("energy")
plot, = pyplot.plot([0],[0])
pyplot.show(block=False)
def update_plot():
    if current_time < 0:
        return
    i = current_time
    plot.set_xdata(energies[:i+1,0])
    plot.set_ydata(energies[:i+1,1])
    pyplot.xlim(0, energies[i,0])
    pyplot.ylim(energies[:i+1,1].min(), energies[:i+1,1].max())
    pyplot.draw()
\end{lstlisting}
Then we merge the \lstinline{main} function definitions from both the previous sections.
\begin{lstlisting}
def main():
    global current_time
    for i in range(0, int_n_times):
        print("run %d at time=%f " % (i, system.time))
        system.integrator.run(int_steps)
        energies[i] = (system.time, system.analysis.energy()['total'])
        current_time = i
        visualizer.update()
        # update_plot() cannot be called from here
\end{lstlisting}
However, as we now have multiple threads, we cannot simply call \lstinline{update_plot()} from the \lstinline{main} function definition.
Instead, we register it as a callback with the visualizer before we start up the visualizer GUI:
\begin{lstlisting}
t = Thread(target=main)
t.daemon = True
t.start()
visualizer.registerCallback(update_plot, interval=500)
visualizer.start()
\end{lstlisting}

\end{document}
