\chapter{\es{} command reference}
\label{chap:ref}

(most from paper and from related pages)

\begin{itemize}
\item Will contain description of all Tcl-commands
\item Reference to appendix \vref{chap:quickref}
\item basic identifiers:
  \begin{itemize}
  \item particle type
  \item bonded interaction type
  \item molecule id
  \end{itemize}
\end{itemize}


\section{\texttt{inter}: Setting up interactions}
\label{sec:inter}
\begin{command}
  \variant{1}
  {inter
    \var{part\_type\_id1} 
    \var{part\_type\_id2}
    \var{inter\_type} 
    [ \var{parameters}\ldots ]
  }
  \variant{2}
  {inter
    \var{bond\_type\_id} \var{bond\_type} [
    \var{parameters}\ldots ]
  }
  \variant{3}
  {inter}
\end{command}

\section{Writing VTF files}
There are two commands in \es{} that support writing files in the VMD
formats VTF, VSF and VCF.\footnote{A description of the format and a
  plugin to read the format in VMD is found in the subdirectory
  \texttt{vmdplugin/} of the \es{} source directory.} The commands can
be used to write the structure (\texttt{writevsf}) and coordinates
(\texttt{writevcf}) of the system to a single trajectory file (usually
with the extension \texttt{.vtf}), or to separate files (extensions
\texttt{.vsf} and \texttt{.vtf}).

\subsection{\texttt{writevsf}}

\begin{command}
  writevsf \var{file} \var{options}
\end{command}
Writes a structure block describing the system's structure to
\var{file}. The atom ids used in the file are identical to \es's
particle ids.  This makes it easy to write additional structure lines
to the file, e.g. to specify the \texttt{resname} of particle
compounds, like chains.  The output of this file can be used in a
standalone VSF file, or at the beginning of a trajectory VTF file that
contains a trajectory of a whole simulation.

\paragraph{Options}
\begin{options}
  short|verbose
\end{options}
Specify, whether the output is in a human-readable, but somewhat
longer format (\keyword{verbose}), or in a more compact form
(\keyword{short}). The default is \keyword{verbose}.

\begin{options}
  radius \var{radii}|auto
\end{options}
Specify the VDW radii of the atoms. \var{radii} is either
\keyword{auto}, or a Tcl-list describing the radii of the different
particle types. When the keyword \keyword{auto} is used and a
Lennard-Jones interaction between two particles of the given type is
defined, the radius is set to be $\frac{\sigma_{LJ}}{2}$ plus the LJ
shift.  Otherwise, the radius $0.5$ is substituted. The default is
\keyword{auto}.  

Example: \verb!radius {{0 2.0} {1 auto} {2 1.0}}!

\begin{options}
  typedesc \var{typedesc}
\end{options}
\var{typedesc} is a Tcl-list giving additional VTF atom-keywords to
specify additional VMD characteristics of the atoms of the given type.
If no description is given for a certain particle type, it defaults to
\texttt{segid \textit{n}}, where \textit{n} is the type id.

Example: \verb!typedesc {{0 "segid colloid"} {1 "segid pe"}}!


\subsection{\texttt{writevcf}}
\begin{command}
  writevcf \var{file} \var{options}
\end{command}
Writes a coordinate (or timestep) block that contains all coordinates
of the system's particles to \var{file}.

\paragraph{Options}
\begin{options}
  short|verbose
\end{options}
Specify, whether the output is in a human-readable, but somewhat
longer format (\keyword{verbose}), or in a more compact form
(\keyword{short}). The default is \keyword{verbose}.

\begin{options}
  folded|absolute
\end{options}
Specify whether the particle positions are written in absolute
coordinates (\keyword{absolute}) or folded into the central image of a
periodic system (\keyword{folded}). The default is \keyword{absolute}.

\begin{options}
  pids \var{pids}|all
\end{options}
Specify the coordinates of which particles should be written. If
\keyword{all} is used, all coordinates will be written (in the ordered
timestep format). Otherwise, \var{pids} has to be a Tcl-list
specifying the pids of the particles. The default is \keyword{all}.

Example: \verb!pids {0 23 42}!



    % \section{Creating particles}
% \label{sec:part}
% \texttt{part}

% \section{Interactions}
% \label{sec:inter}
% \texttt{inter}

% \section{Analysis}
% \label{sec:analysis}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ug"
%%% End: 
