\chapter{Installation}
\label{chap:install}
\index{Installation|textbf}

\begin{itemize}
\item Compiling \es{} is a necessary evil
\item Features can be compiled in or not
\item For maximal efficiency, compile in only the features that you
  use
\item \es{} can be obtained from the \es{} home page
  \footnote{\url{http://www.espresso.mpg.de}}.
\end{itemize}

\section{Requirements}
\label{sec:requirements}
\index{requirements}

\begin{description}
\item[Tcl/Tk] \index{Tcl/Tk} \es{} requires the Toolkit Command
  Language Tcl/Tk \footnote{\url{http://www.tcl.tk/}} in the version
  8.3 or later.  Some example scripts will only work with Tcl 8.4. You
  do not only need the interpreter, but also the header files and
  libraries.  Depending on the operating system, these may come in
  separate development packages. If you want to use a graphical user
  interface (GUI) for your simulation scripts, you will also need Tk.
  
\item[FFTW] \index{FFTW} In addition, \es{} needs the FFTW library
  \footnote{\url{http://www.fftw.org/}} for Fourier transforms.
  ESPResSo can work with both the 2.1.x and 3.0.x series. Again, the
  header files are required.
  
\item[MPI] \index{MPI} Finally, if you want to use \es{} in parallel,
  you need a working MPI environment (version 1.2). Currently, the
  following MPI implementations are supported:
  \begin{itemize}
  \item LAM/MPI is the preferred variant
  \item MPICH, which seems to be considerably slower than LAM/MPI in
    our benchmarks.
  \item On AIX systems, \es{} can also use the native POE parallel
    environment.
  \item On DEC/Compaq/HP OSF/Tru64, \es{} can also use the native
    dmpirun MPI environment.
  \end{itemize}
\end{description}

\section{Quick start}

\index{configure}\index{make}

In many cases, to compile \es{}, it is enough to execute the following
sequence of two steps in the directory where you have unpacked the
sources:
\begin{verbatim}
> configure
> make
\end{verbatim}

In some cases, \eg{} when \es{} needs to be compiled for several
different platforms or when different versions with different sets of
features are required, it might be useful to execute the commands not
in the source directory itself, but to start \texttt{configure} from
another directory (see section \vref{sec:builddir}). Furthermore, many
features of \es{} can be selectively turned on or off in the local
configuration header of \es{} (see section \vref{sec:myconfig}) before
starting the compilation with \texttt{make}.

The shell script \texttt{configure} prepares the source code for
compilation. It will determine how to use and where to find the
different libraries and tools required by the compilation process, and
it will test what compiler flags are to be used.  The script will find
out most of these things automatically.  If something is missing, it
will complain and give hints how to solve the problem.  The
configuration process can be controlled with the help of a number of
options that are explained in section \vref{sec:configure}.

The command \texttt{make} will compile the source code. Depending on
the options passed to the program, \texttt{make} can also be used for
a number of other things:
\begin{itemize}
\item It can install and uninstall the program to some other
  directories. However, normally it is not necessary to actually
  \textit{install} \es{} to run it.
\item It can test the \es{} program for correctness.
\item It can build the documentation.
\end{itemize}
The details of the usage of \texttt{make} are described in section
\vref{sec:make}.

When these steps have successfully completed, \es{} can be started
with the command (see section \vref{sec:run})
\begin{verbatim}
> Espresso
\end{verbatim}

\section{Source and build directory}
\label{sec:builddir}
\index{build directory} \index{source directory}

If you plan to use \es{} with a single configuration, you can skip the
rest of this section. If then you have problems finding the \es{}
binary or you come upon a reference to the \emph{build directory} in
the documentation, you might have to read it, anyway. 

Usually, when a program is compiled, the resulting binary files are
put into the same directory as the sources of the program. In \es{},
the \emph{source directory} that contains all the source files can be
completely separated from the \emph{build directory} where the files
created by the build process are put. As the source directory is not
touched during the compilation process, it is possible to compile more
than one binary version of \es{} from the same set of source files.
This is useful in cases when \es{} is to be used on different computer
hardware or with a different configuration.

The source directory is the directory that contains the source files.
The location of the build directory is determined when the
\texttt{configure}-script is called.  Usually, the build directory is
assigned to the current working directory when the
\texttt{configure}-script was called. All further commands concerning
compiling and running \es{} have to be called from this directory.

\paragraph{Example}
When the source directory is \texttt{\$srcdir} (\ie{} the files where
unpacked to this directory), then the build directory can be set to
\texttt{\$builddir} by calling the \texttt{configure}-script from
there:
\begin{verbatim}
> cd $builddir
> $srcdir/configure
> make
> Espresso
\end{verbatim}

When \texttt{configure} is called directly from the source directory,
the \es{} build system is prepared to handle different platforms.  A
new subdirectory is created and \texttt{configure} is recursively
called from this directory, making the subdirectory the build
directory.  The directory is called
\texttt{obj-}\textit{platform}\texttt{/}, where \textit{platform} is a
descriptor of the CPU type where the script was started, \eg{}
\texttt{obj-Athlon\_64-pc-linux}.

In this case it is also possible to run the commands \texttt{make} and
\texttt{Espresso} directly in the source directory.

Furthermore, the option \texttt{--enable-chooser} will be set in the
recursive call of \texttt{configure} that activates the automatic
binary chooser (see section \vref{sec:install_dir}).

\section{Installation directories}
\label{sec:install_dir}

Normally, the \es-binary \texttt{Espresso-bin} is installed in the
directory \texttt{\$prefix/libexec/} and a the wrapper script
\texttt{Espresso} in the directory \texttt{\$prefix/bin/} that handles
the MPI invocation.

When the \texttt{configure}-script is called from the source directory
or when the option \texttt{--enable-chooser} is given, an automatic
binary chooser is installed in the directory \texttt{\$prefix/bin/}
and the \es{}-binary and the MPI wrapper script are installed in an
architecture-specific subdirectory
\mbox{\texttt{\$exec-prefix/lib/espresso/obj-}\textit{platform}\texttt{/}}.
When called, the binary chooser will automatically call the MPI
wrapper script in the right subdirectory.

\section{The configuration header \texttt{myconfig.h}}
\label{sec:myconfig}

\index{myconfig.h} \index{configuration header} \es{} has a great
number of features that can be compiled into the binary (see chapter
\vref{chap:features}).  However, it is not recommended to actually
compile in all possible features, as this will negatively affect \es's
performance. Instead, compile in only the features that are actually
required. For the developers, it is also possible to turn on or off a
number of debugging messages. The features and debug messages can be
controlled via a configuration header file that contains
C-preprocessor declarations. See Sec.\ref{sec:configflags} for possible
declarations.

By default, the configuration header is called \texttt{myconfig.h}.
The name of the configuration header can be either changed when the
\texttt{configure}-script is called with the option
\texttt{--with-myconfig} (see section \vref{sec:configure}), or when
\texttt{make} is called with the setting
\texttt{myconfig=}\textit{myconfig\_header} (see section
\vref{sec:make}).

The configuration header can be put in the build directory, or in the
source directory. When a configuration header is found in both
directories, the one in the build directory will be used. If both
directories do not contain a configuration header, a default header
will be used that turns on the default features.

The file \texttt{myconfig-sample.h} in the source directory contains
an example configuration header.

\paragraph{Example}
The configuration header can be used to compile different versions
from the same source directory. Suppose that you have a source
directory \texttt{\$srcdir} and two build directories
\texttt{\$builddir1} and \texttt{\$builddir2} that contain different
configuration headers:

\begin{itemize}
\item \texttt{\$builddir1/myconfig.h}:
\begin{verbatim}
#define ELECTROSTATICS
#define LENNARD-JONES
\end{verbatim}

\item \texttt{\$builddir2/myconfig.h}:
\begin{verbatim}
#define LJCOS
\end{verbatim}
\end{itemize}

\noindent Then you can simply compile two different versions of \es{} via
\begin{verbatim}
cd $builddir1
$srcdir/configure
make

cd $builddir2
$srcdir/configure
make
\end{verbatim}

\section{Configuration options}
\label{sec:configflags}
\newcommand\configswitch[1]{\texttt{\bf #1}}

In \texttt{myconfig-sample.h} you can use the following general switches:
\begin{itemize}
\item \configswitch{PARTIAL\_PERIODIC} By default, all coordinates in \es{} are periodic. With
  \texttt{PARTIAL\_PERIODIC} turned on, the \es{} global variable \texttt{periodic} (see
  Sec.~\ref{sec:globalvar}) controls the periodicity of the individual coordinates. Note that this
  slows the integrator down by around $10-30\%$.
\item \configswitch{ELECTROSTATICS} This switches on the various electrostatics algorithms, such as
  the Ewald summation. See Sec.~\ref{sec:electrostatics} for details on this algorithms.
\item \configswitch{ROTATION} Switch on rotational degrees of freedom for the particles, as well as
  the corresponding quaternion integrator. See Sec.~\ref{sec:rotation} for details.
\item \configswitch{DIPOLES} This activates the dipole support in P$^3$M. Currently, a mixing of
  dipoles and charges is not possible, i.~e. all particles have to have charge $q=0$.
  Requires \texttt{ELECTROSTATICS} and \texttt{ROTATION}.
\item \configswitch{EXTERNAL\_FORCES} Allows to define an arbitrary constant force for each particle
  individually. Also allows to fix individual coordinates of particles, e.~g. keep them at a fixed
  position or within a plane.
\item \configswitch{CONSTRAINTS} Turns on various spatial constraints such as spherical compartments
  or walls. This constraints interact with the particles through regular short ranged potentials
  such as the Lennard--Jones potential. See Sec.~\ref{sec:constraints} for possible constraint
  forms.
\item \configswitch{MASS} Allows particles to have individual masses. Note that some analyzation
  procedures have not yet been adapted to take the masses into account correctly.
\item \configswitch{EXCLUSIONS} Allows to exclude specific short ranged interactions within
  molecules, which is necessary for some atomistic models.
\item \configswitch{COMFORCE}
\item \configswitch{COMFIXED}
\item \configswitch{MOLFORCES}
\item \configswitch{BOND\_CONSTRAINT} Turns on the RATTLE integrator which allows for fixed lengths
  bonds between particles.
\end{itemize}

In addition, there are switches that enable additional features in the integrator:
\begin{itemize}
\item \configswitch{NEMD} Enables the non-equilbrium (shear) MD support (see Sec.~\ref{sec:NEMD}).
\item \configswitch{NPT} Enables an on--the--fly NPT integration scheme (see Sec.~\ref{sec:NPT}).
\item \configswitch{DPD} Enables the dissipative particle dynamics thermostat (see
  Sec.~\ref{sec:DPD}).
\item \configswitch{LB} Enables the lattice-Boltzmann fluid code (see Sec.~\ref{sec:LB}).
\end{itemize}

\subsection{Switches for interactions}
The following switches turn on various short ranged interactions (see Sec.~\ref{sec:shortrange}):
\begin{itemize}
\item \configswitch{TABULATED} Enable support for user--defined interactions, e.~g. for atomistic
  simulations.
\item \configswitch{LENNARD\_JONES} Enable the Lennard--Jones potential.
\item \configswitch{LJ\_WARN\_WHEN\_CLOSE} This adds an additional check to the Lennard--Jones
  potential that prints a warning of particles come too close so that the simulation becomes
  unphysical.
\item \configswitch{MORSE} Enable the Morse potential.
\item \configswitch{LJCOS} Enable the Lennard--Jones potential with a cosine--tail.
\item \configswitch{LJCOS2}
\item \configswitch{BUCKINGHAM} Enable the Buckingham potential.
\item \configswitch{SOFT\_SPHERE} Enable the soft sphere potential.
\end{itemize}

If you want to use angle bonds, you currently need to choose the type a priory (see
Sec.~\ref{sec:angle}). This will change in the near future to three independent angle potentials:
\begin{itemize}
\item \item \configswitch{BOND\_ANGLE\_HARMONIC}
\item \configswitch{BOND\_ANGLE\_COSINE}
\item \configswitch{BOND\_ANGLE\_COSSQUARE}
\end{itemize}

\subsection{Debug--switches}
Finally, there are a number of flags for debugging. The most important one are
\begin{itemize}
\item \configswitch{ADDITIONAL\_CHECKS} Enables numerous additional checks which can detect
  inconsistencies especially in the cell systems. This checks are however too slow to be enabled in
  production runs.
\item \configswitch{MEM\_DEBUG} Enables an internal memory allocation checking system. This produces
  output for each allocation and freeing of a memory chunk, and therefore allows to track down
  memory leaks. This works by internally replacing \texttt{malloc}, \texttt{realloc} and
  \texttt{free}.
\end{itemize}

The following flags control the debug output of various sections of Espresso. You will however
understand the output very often only by looking directly at the code.
\begin{itemize}
\item \configswitch{COMM\_DEBUG} Output from the asynchronous communication code.
\item \configswitch{EVENT\_DEBUG} Notifications for event calls, i.~e. the \texttt{on\_?} functions
  in \texttt{initialize.c}. Useful if some module does not correctly respond to changes of e.~g.
  global variables.
\item \configswitch{INTEG\_DEBUG} Integrator output.
\item \configswitch{CELL\_DEBUG} Cellsystem output.
\item \configswitch{GHOST\_DEBUG} Cellsystem output specific to the handling of ghost cells and the
  ghost cell communication.
\item \configswitch{GHOST\_FORCE\_DEBUG}
\item \configswitch{VERLET\_DEBUG} Debugging of the Verlet list code of the domain decomposition cell
  system.
\item \configswitch{LATTICE\_DEBUG} Universal lattice structure debugging.
\item \configswitch{HALO\_DEBUG}
\item \configswitch{GRID\_DEBUG}
\item \configswitch{PARTICLE\_DEBUG} Output from the particle handling code.
\item \configswitch{P3M\_DEBUG}
\item \configswitch{ESR\_DEBUG} debugging of P$^3$Ms real space part.
\item \configswitch{ESK\_DEBUG} debugging of P$^3$Ms $k$--space part.
\item \configswitch{EWALD\_DEBUG}
\item \configswitch{FFT\_DEBUG} Output from the unified FFT code.
\item \configswitch{MAGGS\_DEBUG}
\item \configswitch{RANDOM\_DEBUG}
\item \configswitch{FORCE\_DEBUG} Output from the force calculation loops.
\item \configswitch{THERMO\_DEBUG} Output from the thermostats.
\item \configswitch{LJ\_DEBUG} Output from the Lennard--Jones code.
\item \configswitch{MORSE\_DEBUG} Output from the Morse code.
\item \configswitch{FENE\_DEBUG}
\item \configswitch{ONEPART\_DEBUG} Define to a number of a particle to obtain output on the forces
  calculated for this particle.
\item \configswitch{STAT\_DEBUG}
\item \configswitch{POLY\_DEBUG}
\item \configswitch{MOLFORCES\_DEBUG}
\item \configswitch{LB\_DEBUG} Output from the lattice--Boltzmann code.
\item \configswitch{ASYNC\_BARRIER} Introduce a barrier after each asynchronous command
  completion. Helps in detection of mismatching communication.
\item \configswitch{FORCE\_CORE} Causes \es{} to try to provoke a core dump when exiting
  unexpectedly.
\item \configswitch{MPI\_CORE} Causes \es{} to try this even with MPI errors.
\end{itemize}

\section{Running configure}
\label{sec:configure}

\index{configure}
The shell script \texttt{configure} collects all the information
required by the compilation process. It will determine how to use and
where to find the different libraries and tools required by the
compilation process, and it will test what compiler flags are to be
used.  The script will find out most of these things automatically.
If something is missing, it will complain and give hints how to solve
the problem.

The generic syntax of calling the \texttt{configure} script is:
\begin{syntax}
 $>$ configure [\var{options} ...] [\var{variable}=\var{value} ...]
\end{syntax}

\index{configure options}
The behaviour of \texttt{configure} can be controlled by the means of
command line options. In the following, only those command line
options that are special to \es{} will be explained.  For a complete
list of options and explanations thereof, call
\begin{verbatim}
> configure --help
\end{verbatim}

\begin{description}
\item [\texttt{--enable-chooser}] This option will enable the
  automatic choosing mechanism for \es{} (see section
  \vref{sec:install_dir}).  This option will be automatically enabled,
  when the \texttt{configure} script is called from the source
  directory, otherwise it will be disabled. It is not recommended to
  set the option manually.
\item[\texttt{--enable-config=KNOWN\_CONFIG}] For some known systems,
  where \texttt{configure} does not find the required libraries and
  compiler options, predefined settings can be used. The following
  configuration names are known: \texttt{dino} and \texttt{blade}. The
  default for this option is: \texttt{none}.
\item[\texttt{--enable-debug}] This option will enable compiler flags
  required for debugging \es{} and is disabled by default.
\item[\texttt{--enable-profiling}] This option will enable compiler
  flags required for profiling \es{} and is disabled by default.
\item[\texttt{--disable-processor-optimization}] This option will
  control whether \texttt{configure} will check for several
  optimization flags to be used by the compiler. This option is
  enabled by default.
\item[\texttt{--enable-xlc-qipa}] This option is only useful when the
  IBM C-compiler \texttt{xlc} is used and will control whether or not
  the compiler flag \texttt{-qipa} is used. This option is enabled by
  default.

\item[\texttt{--with-myconfig=MYCONFIG\_HEADER}] This option sets the
  name of the local configuration header (see \vref{sec:myconfig}). It
  defaults to ``\texttt{myconfig.h}''.
\item[\texttt{--with-mpi=MPI}] Sets the MPI implementation that should
  be used. By default, \texttt{configure} will test autoamtically what
  MPI implementation is available. The following implementations are
  known: 
  \begin{description}
  \item[\texttt{fake}, \texttt{no}] This will disable MPI completely.
  \item[\texttt{lam}] Use the LAM/MPI environment
    (\url{http://www.lam-mpi.org/}).
  \item[\texttt{mpich}] Use the MPICH environment
    (\url{http://www-unix.mcs.anl.gov/mpi/mpich/}).
  \item[\texttt{poe}] Use the POE environment (IBM).
  \item[\texttt{dmpi}] Use the DMPI environment (Tru64).
  \item[\texttt{generic}] Use a generic MPI implementation. This will
    try to find an MPI compiler and an MPI runtime environment.
  \end{description}
\item[\texttt{--with-efence}] Whether or not to use the ``electric
  fence'' memory debugging library
  (\url{http://freshmeat.net/projects/efence/}). Efence is not used by
  default.
\item[\texttt{--with-tcl=TCL}] When the wrong version of the Tcl
  library is used by configure, the name of the Tcl-library can be
  specified with this option, \eg{} \texttt{tcl8.4}.
\item[\texttt{--with-tk=TK}] By default, the GUI toolkit Tk is not
  used by \es. This option can be used to activate Tk and to specify
  which Tk version to use, \eg{} \texttt{tk8.4}.
\item[\texttt{--with-fftw=VERSION}] This can be used to specify which
  version of fftw is to be used. By default, version 3 will be used if
  it is found, otherwise version 2 is used.
\end{description}

\section{Compiling, testing and installing \es}
\label{sec:make}

The command \texttt{make} is mainly used to compile the \es{} source
code, but it can do a number of other things. The generic syntax of
the \texttt{make} command is:
\begin{syntax}
 $>$ make [\var{target}...] [\var{variable}=\var{value}]
\end{syntax}
When no target is given, the target \texttt{all} is used. The
following targets are available:
\begin{description}
\item[\texttt{all}] Compiles the complete \es{} source code.
\item[\texttt{check}] Runs the testsuite. By default, all available
  tests will be run on 1, 2, 3, 4, 6, or 8 processors. Which tests are
  run can be controlled by means of the variable \texttt{tests}, which
  processor numbers are to be used can be controlled via the variable
  \texttt{processors}. Note that depending on your MPI installation,
  MPI jobs can only be run in the queueing system, so that \es{} will
  not run from the command line. In that case, you may not be able to
  run the testsuite, or you have to directly submit the testsuite script
  \verb!testsuite/test.sh! to the queueing system.\\
  \textbf{Example:} \verb!make check tests="madelung.tcl" processors="1 2"!\\
  will run the test \texttt{madlung.tcl} on one and two processors.
\item[\texttt{clean}] Deletes all files that were created furing the
  compilation.
\item[\texttt{mostlyclean}] Deletes most files that were created
  during the compilation. Will keep for example the built doxygen
  documentation and the \es{} binary.
\item[\texttt{dist}] Creates a \texttt{.tar.gz}-file of the \es{}
  sources.  This will include all source files as they currently are
  in the source directory, \ie{} it will include local changes.  This
  is useful to give your version of \es{} to other people.
  The variable \texttt{extra} can be used to specify additional
  files and directories that are to be included in the archive
  file. \\
  \textbf{Example:} \verb!make dist extra="myconfig.h internal"!\\
  will create the archive file and include the file
  \texttt{myconfig.h} and the directory \texttt{internal} with all
  files and subdirectories.
\item[\texttt{install}] Install \es{}. The variables \texttt{prefix}
  and \texttt{exec-prefix} can be used to specify the installation
  directories, otherwise the defaults defined by the
  \texttt{configure} script are used. \texttt{prefix} sets the basic
  prefix where all \es{} files are to be installed,
  \texttt{exec-prefix} sets the prefix where the executable files are
  to be installed and is required only when there is an
  architecture-specific directory.\\
  \textbf{Example:} \verb!make install prefix=/usr/local!\\
  will install all files below \texttt{/usr/local}.
\item[\texttt{uninstall}] Uninstalls \es{}, \ie{} removes all files
  that were installed during \texttt{make install}. The variables are
  identical to the variables of the \texttt{install}-target.
\end{description}

\section{Running \es}
\label{sec:run}

\es{} can be run via
\begin{syntax}
$>$ Espresso [\var{tcl\_script} [\var{N\_processors} [\var{args}]]]
\end{syntax}

\index{interactive mode} When \es{} is called without any arguments,
it is started in the interactive mode, where new commands can be
entered on the command line. When the name of a \textit{tcl\_script}
is given, the script is executed. \textit{N\_processors} is the number
of processors that are to be used. Any further arguments are passed to
the script. Note that depending on your MPI installation, MPI jobs can
only be run in the queueing system, so that \es{} will not run from
the command line.

% A number of wrapper scripts are used in running \es{}:
% \begin{itemize}
% \item The script \texttt{Espresso} in the source and build directory
%   will try to run the compiled version of \es. If it is called from
%   the source directory, it assumes that \es{} was also configured in
%   the source directory and will try to recursively start the script in
%   the corresponding \texttt{obj-PLATFORM} build directory. If it is
%   called in the build directory, it will start the \es-binary with the
%   right MPI implementation.
% \item The chooser script \texttt{Espresso} 
%   \begin{itemize}
%   \item installed when \verb!--enable-chooser! was given
%   \item installed to bindir
%   \item tries to run the correct version of the MPI-wrapper
%     \texttt{Espresso}
%   \end{itemize}
% \item The MPI-wrapper \texttt{Espresso}
%   \begin{itemize}
%   \item installed next to \es{} binary
%   \item starts the binary with the right MPI implementation
%   \end{itemize}
% \item The \es{} binary \texttt{Espresso-bin} can also be started
%   directly, however, it requires that the environment variable
%   \verb!ESPRESSO_SCRIPTS! is set to the directory where the scripts
%   are installed (usually \verb!$(prefix)/lib/espresso/scripts! or
%   \verb!$(prefix)/share/espresso/scripts!).
% \end{itemize}



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ug"
%%% End: 
